Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-09 text/back_end_KS.qmd
working through back end

```{=tex}
\pagenumbering{roman}
\newpage
```
**ABSTRACT**

@@

```{=tex}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```
# Data and Methods

@@

**RQ1**. Our study utilizes two main data sources based on each research question. The first research question is focused on understanding the socioeconomic and racial variation between Geomarkets in metropolitan areas and how such variation changes over time. To answer this question, we use census tract data from the U.S. Census Bureau. Utilizing a small geographic area is important for implementing a spatial merge between Census data and Geomarket shape files (described below). Census tract is the smallest geographic area for which measures of both race/ethnicity and socioeconomic characteristics are available. Census tracts are statistical subdivisions of a county designed to capture relatively homogeneous demographic communities that are generally considered to be more accurate for population analyses than zip codes, which are spatial boundaries drawn to demarcate route efficiency for postal services \[[CITE](https://www.census.gov/newsroom/blogs/random-samplings/2014/07/understanding-geographic-relationships-counties-places-tracts-and-more.html)\] .

For the first research question, we use data from the 1980 and 2000 Decennial Census as well as 5-year estimates from the 2020 American Community Survey (ACS). For the 1980 Decennial Census, we specifically draw on Summary Tape File 1 (STF1) for variables about race and Hispanic origin collected via “short form” questions answered by all households and Summary Tape File 3 (STF3) for variables about socioeconomic characteristics (income, education, and poverty) collected via the "long form" questionnaire completed by a sample of households. The 1980 Decennial Census data collection period closely matches the data collection period of SAT score-sending behavior that @RN4982 utilized to create Geomarkets and the Market Segment Model. To examine how Geomarket characteristics change over time, we similarly use the Summary File 1A (SF1A) for census-tract measures of race and Hispanic origin and the Summary File 3A (SF3A) for census-tract measures of income, education, and poverty from the 2000 Decennial Census.[^1] Beginning in 2010, the Decennial Census no longer collected a "long form" questionnaire. These questions were replaced by ACS. Therefore, we also utilized the 2020 5-year ACS for both race and socioeconomic characteristics at the census tract level, which includes data collected from 2016-2020.[^2] This data collection period mirrors the period for our primary data collection about student list purchases used to analyze RQ2, as described below.

@@

Across these data sources, we created measures of race and ethnicity, specifically the number and proportion of the census-tract population in the following categories: White, non-Hispanic; Black, non-Hispanic; Asian, non-Hispanic; Pacific Islander, non-Hispanic; American Indian and Alaskan Native, non-Hispanic; and two or more races, non-Hispanic. Measures of Asian, American Indian and Alaskan Native, two or more races, and Native Hawaiian and Pacific Islander (all non-Hispanic) were not available for the 1980 Decennial Census. We also created socioeconomic measures that approximate the predictors of market segment identified by @RN4982. These variables are median household income, percent of households below the poverty line, and people age 25+ with a BA. All measures were created at the census tract-level and the Geomarket-level by aggregating tract-level data to Geomarkets.

@@

Next, we utilized the `st_intersection()` function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. Census tract shapefiles that intersected with multiple Geomarkets (i.e., a census tract's spatial boundaries were spread across two or more Geomarkets) broken into smaller shapefiles wholly contained within a single Geomarket. The variables in these shapefiles were then replaced with the original variable value multiplied by the proportion of land area of the census tract that was contained within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 1,000 people who identify as Hispanic non-white. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*1000 = 600$ Hispanic non-white people and the observation assigned to Geomarket B reports $.40*1000 = 400$ Hispanic non-white people.

**RQ2**. Our study's second research question asks how does the racial and socioeconomic composition of included versus included prospects vary when student list purchases filter on Geomarkets. We answer this question using actual student lists purchased by public universities. In February 2020, we began issuing public records requests to public universities about student lists purchased from 2016 through 2020. We narrowed the scope of our request to student lists purchased from College Board, ACT, and the National Research Center for College and University Admissions (NRCCUA), the three largest student list vendors at the time. For each student list purchased from 2016 through 2020, we requested two related pieces of data: (1) the order summary, which specifies search criteria for the student list purchase; and (2) the de-identified prospect-level list produced from the search criteria.

The data collection sample for this project was all public universities in California, Illinois, Minnesota, and Texas. Utilizing public records requests to obtain public records is a painstaking process. Initially, the majority of universities did not respond or denied our request. Subsequently, we obtained pro bono representation from four law firms, which substantially increased the success of data collection. We also collected data from Arizona State University and Northern Arizona University, because we were able to obtain pro-bono legal representation for these two universities. However, we were unable to obtain legal representation for Texas.

Even with firm representation, data collection remained difficult. Some universities provided records that were not usable for quantitative analyses (e.g., summary statistics across multiple orders; or data did not contain important fields). Some universities did not provide records based on legitimate legal grounds (e.g., data not in university possession; not required to create records that do not currently exist). We learned that many universities outsourced student list purchases to third-party consulting firms. Unfortunately, we were rarely able to obtain usable data from these universities. A small number of universities denied requests based on potentially questionable legal rationale, but we lacked the resources to litigate.

The primary data source for our second research question is therefore student lists purchased from College Board for which we received both (a) the order summary data and (b) the de-identified prospect-level student list data. This sample includes a total of 414 orders associated with 2,549,085 prospects made by 14 \[KS CHECK THIS\] public universities located in California, Illinois, Minnesota, and Arizona \[KS Does this still apply?\]. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase.[^4] The resulting student lists could hypothetically extend to any university placing student lists orders based on the same filter criteria. In other words, any observed patterns in the findings below would remain constant for any university selecting the same filters. Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

@@

In particular, we answer the second research question by analyzing particular student list purchases that did *not* utilize the Geomarket search filter, but did purchase prospects from all Geomarkets in a particular metropolitan area. For each prospect we know their home zip code and, therefore, which Geomarket they belong to. We therefore simulate which prospective students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets.

The variables of interest for student list data are the characteristics of prospects who would have been included versus excluded by our simulations by Geomarket filter. Prospect characteristics are derived from the pretest questionnaire administered to College Board test-takers that are included in the de-identified student lists we received (e.g., HS code, HS GPA range, intended major). College Board asks students separate questions about ethnicity (Cuban, Mexican, Puerto Rican, other Hispanic, non-Hispanic, ethnicity nonresponse) and race (American Indian or Alaska Native, Asian, Black, Native Hawaiian or other Pacific Islander, White, race non-response), allowing students to check as many boxes as they want (College Board, 2016). From these responses, a College Board "derived aggregate race/ethnicity" was re-created based on the U.S. Department of Education reporting guidelines that includes the following categories: no response; American Indian/Alaska Native; Asian; Black; Hispanic/Latino; Native Hawaiian or Other Pacific Islander; White; other; two or more races, non-Hispanic (College Board, 2016).[^5] Information about student socioeconomic status is limited. However, we create a measure of whether the prospect is first-generation college student based on their parents' level of education that was self-reported on the College Board questionnaire \[KS double check this\].

@@

**Research design**. We utilize a multiple, quantitative case study design in which metropolitan areas are cases. Case studies are used to gain a deeper understanding of the phenomenon of interest by comparing and contrasting across different settings \[CITE YIN 2001\]. Our study's research questions explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to generalize Geomarket variation nationally, but rather to explore how differences across Geomarket landscapes shape prospect pools based on the geographical model of metropolitan areas and their demographic characteristics. Such aims align with the purpose of case studies, which drive the exploration of research questions focused on the "how" and "why" of phenomenon in rich and nuanced contexts. We therefore focus on minimizing the number of metropolitan areas in order to provide sufficient analytic depth while situating each case within its specific historic and contemporary sociodemographic characteristics \[TIE IN THEORETICAL FRAMEWORK HERE A BIT MORE\]. Online Appendix XXX presents results for a larger number of metropolitan areas.

Case selection was therefore informed by purposeful sampling to identify “information-rich” cases (Patton, 2002; Yin, 2009) that can provide in-depth knowledge about variation across Geomarkets and overtime. We considered metropolitan areas as information rich cases across three broad dimensions: 1) urban model structures, 2) geographical variation, 3) variation across racial/ethnic and socioeconomic characteristics, and 4) availability of student list orders and de-identified list data. Based on our read of @RN4982, we identified classic urban model structures that reflected "major metropolitan area\[s\] \ldots composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs" (p. 11-12), such as Boston, Chicago, Detroit, Baltimore, Denver, Houston, Miami, Philadelphia**,** Phoenix, and Pittsburgh. To compare and contrast with classic urban model structures, we also identified alternative urban models such as bimodal or metroplex models that contain two centralized cores surrounded by the more classing suburban ring (e.g., Dallas-Fort Worth and San Francisco) and multi-nuclei models where urban growth develops around multiple, sometimes uncentralized and independent, metropolitan areas (e.g., Los Angeles and New York City). The third consideration addresses geographical variations in demographic populations by intentionally selecting metropolitan cases across different U.S. regions (West, Midwest, Northeast, South) to ensure resident and prospect pools mirror broader populations. Cumulatively related to all previous selection dimensions, we are also interested in variation of racial and socioeconomic populations between Geomarkets within a metropolitan areas (RQ1) and how such variation shapes which prospective students are targeted by student list purchases (RQ2). We therefore considered whether or not metropolitan areas provided sufficient and varied racial and socioeconomic diversity for our research purposes.

Lastly, choice of metropolitan cases was informed by student list data availability. Although Census data are available for all metropolitan areas nationally (RQ1), we do not have good candidate student list data for all metropolitan areas (RQ2). In order to simulate which prospective students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets, student list data must contain orders that resulted in a large pool of prospects within the metropolitan areas considered above that would include all Geomarkets within metropolitan areas. This lead us to select on orders that filtered for an entire state (e.g., State filter is set to Pennsylvania), an entire metropolitan area without a Geomarket filter (CBSA filter is set to the Philadelphia metropolitan area) , or an entire metropolitan area with a Geomarket filter inclusive of all Geomarkets (CBSA filter is set to the Philadelphia metropolitan area and Geomarket filter is set to all five Geomarkets in the Philadelphia area). In order to isolate the impact of Geomarkets on the racial/ethnic and socioeconomic characteristics of prospect pools, a suitable metropolitan case candidate would also be contingent on student list orders that did not include too many filtering criteria beyond those detailed previously. Because nearly all orders we collected (Author, BLINDED) filter on some academic criteria, we only considered orders that included assessment threshold filters (e.g., PSAT, SAT) across a wide score range to explore whether and to what extent Geomarket contributions to prospect pools are consistent across all score ranges.

@@

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. The maps in the figure show how Chicago Geomarkets reflect the classic urban geographical structure of an inner-city core surrounded by a ring of suburbs as described by @RN4982 in the development of The Market Segment Model. Based on the color gradient scale representing total population density, @fig-chicago-maps indicates northern suburb Geomarkets in the metropolitan area - Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie - experienced relatively little population growth from 1980 through 2020. In contrast, the most significant population increases occured in the Western Suburbs (1.2 Million to 1.6 Million) and the South and Southwest Suburbs (less than 1 Million to 1.3 Million) Geomarkets. The City of Chicago Geomarket, forming the metropolitan area's central core, was the only locality to experience a population decline, from 3.1 Million in 1980 to 2.8 Million in 2020, despite remaining the most densely populated Geomarket.

These population shifts have contributed to changes in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarkets across the three census periods. In 1980, nearly all Geomarkets were predominantly White, with proportions ranging from 83% in the South and Southwest Suburbs to approximately 93% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 46% of its total population. By 2020, several Geomarkets that were previously predominantly White shifted to a more balanced racial/ethnic distribution, including Chain of Lakes (59% White, 8% Black, 24% Hispanic; 7% Asian Pacific Islander), Northwest Suburbs (62% White, 3% Black, 18% Hispanic; 15% Asian Pacific Islander), Evanston and Skokie (64% White, 7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (56% White, 9% Black, 23% Hispanic; 9% Asian Pacific Islander), and South and Southwest Suburbs (53% White, 24% Black, 18% Hispanic; 3% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 19 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie).

@fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area.

@@

From 1980 to 2020, these demographic shifts have transformed the Chicago metropolitan area's Geomarket landscape from largely a binary structure into a more tiered dynamic shaped by racial/ethnic and socioeconomic characteristics. The original Geomarket boundaries delineated significant racial and income segregation in the 1980s. This included Black, Hispanic, and Asian residents primarily in the City of Chicago Geomarket and White residents in Geomarkets forming the suburban ring of the metropolitan area. Income, poverty, and educational attainment also created a static socioeconomic divide in 1980, with southern Geormarkets (City of Chicago and South and Southwest Suburbs) characterized by lower affluence and educational attainment compared to the higher socioeconomic standing of the northwestern suburbs. However, declining White populations, Hispanic and Asian population growth, and Black outmigration, as well as some narrowing of education and poverty gaps, have contributed to a more stratified Geomarket dynamic in the metropolitan area despite significant disparities remaining. The City of Chicago Geomarket remains the critical geographical hub across all Geomarkets for racial/ethnic diversity; however, other Geomarkets include substantive numbers of residents across racial/ethnic groups (such as South and Southwest Geomarket for Black residents, Chain of Lakes for Hispanic residents, and Northwest Suburbs for Asian residents) while the the Evanston and North Shore Geomarkets remain predominantly White. While some stark socioeconomic gaps within Geomarkets have narrowed since 1980, significant geographical income segregation remains across Geomarkets despite a more clearly defined middle-to-high socioeconomic Geomarket group (Chain of Lakes, Northwest Suburbs, Western Suburbs).

@@

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets, we analyze six orders placed by a research university targeting prospects across the entire state of Illinois. Two of these orders applied the same filters: SAT score thresholds ranging from 1020 to 1150, GPA ranging from B- to A+, and high school class targeting students from the 2019-2020 and 2020-2021 graduating classes, respectively. The remaining four orders used similar GPA and high school class filters but differed in SAT thresholds- two targeted students with relatively middle-range scores (1160 to 1300) and two targeted those with relatively high scores (1310 to 1600). Because all six orders included an Illinois state filter, we can utilize prospects from the Chicago metropolitan area within the resulting student lists to examine Geomarket contributions by simulating the application of a Geomarket filter across low-, middle-, and high-range SAT threshold orders.

>>>>

\pagenumbering{roman}
\newpage

**ABSTRACT**

@@

\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}

# Data and Methods

@@

**RQ1**. RQ1 primarily utilizes Census Data while RQ2 primarily utilizes data about student list purchases. Research question 1 is about socioeconomic and racial variation between Geomarkets in metropolitan areas and how such variation changes over time. To answer this question, we use census tract data from the U.S. Census Bureau. Utilizing a small geographic area is important for implementing a spatial merge between Census data and Geomarket shape files (described below). Census tract is the smallest geographic area for which measures of both race/ethnicity and socioeconomic characteristics are available. Census tracts are statistical subdivisions of a county designed to capture relatively homogeneous demographic communities that are generally considered to be more accurate for population analyses than zip codes, which are spatial boundaries drawn to demarcate route efficiency for postal services \[[CITE](https://www.census.gov/newsroom/blogs/random-samplings/2014/07/understanding-geographic-relationships-counties-places-tracts-and-more.html)\] .

We use data from the 1980 and the 2000 Decennial Census as well as 5-year estimates from the 2020 American Community Survey (ACS). For the 1980 Decennial Census, we draw from Summary Tape File 1 (STF1) for variables about race and Hispanic origin collected via "short form" questions answered by all households. We draw from Summary Tape File 3 (STF3) for variables about socioeconomic characteristics (income, education, and poverty) collected via the "long form" questionnaire completed by a sample of roughly 20% of all households. The 1980 Decennial Census data collection period closely matches the data collection period of SAT score-sending behavior that @RN4982 utilized to create Geomarkets and the Market Segment Model. For the 2000 Decennial Census, we similarly use the Summary File 1A (SF1A) for census-tract measures of race and Hispanic origin and the Summary File 3A (SF3A) for census-tract measures of income, education, and poverty.[^1] Beginning in 2010, the Decennial Census no longer collected a "long form" questionnaire. These questions were replaced by ACS. Therefore, we also utilized the 2020 5-year ACS for both race and socioeconomic characteristics at the census tract level, which includes data collected from 2016-2020.[^2] This data collection period mirrors the period for our primary data collection about student list purchases used to analyze RQ2, as described below.

@@

Across these data sources, we created measures of race and ethnicity, specifically the number and proportion of the census-tract population in the following categories: White, non-Hispanic; Black, non-Hispanic; Asian, non-Hispanic; Pacific Islander, non-Hispanic; American Indian and Alaskan Native, non-Hispanic; and two or more races, non-Hispanic. Measures of Asian, American Indian and Alaskan Native, two or more races, and Native Hawaiian and Pacific Islander (all non-Hispanic) were not available for the 1980 Decennial Census. We also created socioeconomic measures that approximate the predictors of market segment identified by @RN4982. These variables are median household income, percent of households below the poverty line, and people age 25+ with a BA. All measures were created at the census tract-level and at the the Geomarket-level by aggregating tract-level data to Geomarkets.

@@

Next, we utilized the `st_intersection()` function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. We split census-tract shapefiles that intersected multiple Geomarkets (i.e., tracts whose boundaries spanned two or more Geomarkets) into smaller shapefiles, each entirely contained within a single Geomarket. We replaced each variable in these shapefiles with its original value scaled by the share of the tract’s land area that lay within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 100 people who identify as Asian, non-Hispanic. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*100 = 60$ Asian, non-Hispanic people and the observation assigned to Geomarket B reports $.40*100 = 40$ Asian, non-Hispanic people.

**RQ2**. Our second research question asks, how does the racial and socioeconomic composition of included versus excluded prospects vary when student list purchases filter on Geomarkets? We answer this question using actual student lists purchased by public universities. In February 2020, we began issuing public records requests to public universities about student lists purchased from 2016 through 2020. We narrowed the scope of our request to student lists purchased from College Board, ACT, and the National Research Center for College and University Admissions (NRCCUA), the three largest student list vendors at the time. For each student list purchased from 2016 through 2020, we requested two related pieces of data: (1) the order summary, which specifies search criteria for the student list purchase; and (2) the de-identified prospect-level list produced from these search criteria.

The data collection sample for this project was all public universities in California, Illinois, Minnesota, and Texas. We chose these states because they were priorities for the two foundations that funded the data collection. Utilizing public records requests to obtain quantitative data is a painstaking process, made more difficult by the onset of the Covid Pandemic. Initially, the majority of universities ignored or denied our request. We established a client relationship with a civil rights legal organization and leveraged this relationship to obtain pro bono representation from four law firms. Firm representation substantially increased the success of data collection. We also collected data from Arizona State University and Northern Arizona University, because we were able to obtain pro-bono legal representation for these two universities. However, we were unable to obtain legal representation for Texas because firms interested in representing us had conflicts of interest in Texas.

Firm representation substantially increased the success of data collection. Nevertheless, data collection remained difficult. Some universities provided records that were not usable for quantitative analyses (e.g., summary statistics across multiple orders; or data did not contain important fields). Some universities did not provide records based on legitimate legal grounds (e.g., data not in university possession; not required to create records that do not currently exist). We learned that many universities outsourced student list purchases to third-party consulting firms, particularly [EAB](https://eab.com/). Unfortunately, we were rarely able to obtain usable data from these universities. Several universities denied requests based on questionable legal rationale, but we lacked the resources to litigate.

The primary data source for our second research question is student lists purchased from College Board for which we received both (a) the order summary data and (b) the de-identified prospect-level student list data. This sample includes a total of 414 orders associated with 2,549,085 prospects made by 14 \[KS CHECK THIS\] public universities located in California, Illinois, Texas, and Arizona. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase. The resulting student lists could hypothetically extend to any university placing student lists orders based on the same filter criteria. In other words, any observed patterns in the findings below would remain constant for any university selecting the same filters.[^4] Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

@@

In particular, we answer the second research question by analyzing particular student list purchases that did *not* utilize the Geomarket search filter and purchased prospects from all Geomarkets in a particular metropolitan area.^[For analyses of Dallas, Detroit, Long Island, Northern New Jersey, and Philadelphia, we utilize data from student list purchases that did filter on Geomarkets but included all Geomarkets in the region.] For each prospect, we know their home zip code and, therefore, which Geomarket they belong to. Our analyses then show which students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets.

The variables of interest are Geomarket and prospect-level characteristics, which are derived from the pretest questionnaire administered to College Board test-takers that are included in the de-identified student lists we received (e.g., HS code, HS GPA range, intended major). College Board asks students separate questions about ethnicity (Cuban, Mexican, Puerto Rican, other Hispanic, non-Hispanic, ethnicity non-response) and race (American Indian or Alaska Native, Asian, Black, Native Hawaiian or other Pacific Islander, White, race non-response), allowing students to check as many boxes as they want (College Board, 2016) [CITE]. From these responses, a College Board "derived aggregate race/ethnicity" was re-created based on the U.S. Department of Education reporting guidelines that includes the following categories: no response; American Indian/Alaska Native; Asian; Black; Hispanic/Latino; Native Hawaiian or Other Pacific Islander; White; other; two or more races, non-Hispanic (College Board, 2016) [CITE].[^5] Information about student socioeconomic status is limited. However, we create a measure of whether the prospect is first-generation college student based on their parents' level of education that was self-reported on the College Board questionnaire \[KS double check this\].

@@

**Research design**. We utilize a multiple, quantitative case study design. Metropolitan areas are cases. Case studies deepen understanding of phenomena by comparing and contrasting across different settings \[CITE YIN 2001\]. Our research questions explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to generalize Geomarket variation nationally, but rather to explore how differences across Geomarket landscapes shape prospect pools based on the geographical model of metropolitan areas and their demographic characteristics [MAKE THIS MORE ABOUT EXPLORING THE MODEL. THIS SENTENCE SOUNDS FANCY BUT DOESN'T REALLY SAY ANYTHING]. Such aims align with the purpose of case studies, which drive the exploration of research questions focused on the "how" and "why" of phenomenon in rich and nuanced contexts. We, therefore, select a small number of metropolitan areas in order to provide sufficient analytic depth while situating each case within its specific historic and contemporary sociodemographic characteristics \[TIE IN THEORETICAL FRAMEWORK HERE A BIT MORE\]. Online Appendix XXX presents results for a larger number of metropolitan areas.

Case selection was informed by purposeful sampling to identify "information-rich" cases (Patton, 2002; Yin, 2009)[CITE] that can provide knowledge about variation across Geomarkets and over time. We considered metropolitan areas as information rich cases across three broad dimensions: 1) urban model structures, 2) geographical variation, 3) variation across racial/ethnic and socioeconomic characteristics, and 4) availability of student list orders and de-identified list data. Based on @RN4982, we identified classic urban model structures that reflected "major metropolitan area\[s\] \ldots composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs" (p. 11-12), such as Boston, Chicago, Detroit, Baltimore, Denver, Houston, Miami, and Philadelphia. To compare and contrast with classic urban model structures, we also identified alternative urban models such as bimodal or metroplex models that contain two centralized cores surrounded by the more classing suburban ring (e.g., Dallas-Fort Worth and San Francisco) and multi-nuclei models where urban growth develops around multiple, sometimes uncentralized and independent, metropolitan areas (e.g., Los Angeles and New York City). The third consideration addresses geographical variations in demographic populations by intentionally selecting metropolitan cases across different U.S. regions (West, Midwest, Northeast, South) to ensure resident and prospect pools mirror broader populations. Related to all previous selection dimensions, we are also interested in variation of racial and socioeconomic populations between Geomarkets within a metropolitan areas (RQ1) and how such variation shapes which prospective students are targeted by student list purchases (RQ2). We therefore considered whether or not metropolitan areas provided sufficient and varied racial and socioeconomic diversity for our research purposes.

Lastly, choice of metropolitan cases was informed by student list data availability. Although Census data are available for all metropolitan areas nationally (RQ1), we do not have good candidate student list data for all metropolitan areas (RQ2). In order to show which prospective students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets, the student list purchases we analyze must include all Geomarkets in a metropolitan area. This lead us to select on orders that filtered for an entire state (e.g., State filter is set to Pennsylvania), an entire metropolitan area without a Geomarket filter (CBSA filter is set to the Philadelphia metropolitan area), or an entire metropolitan area with a Geomarket filter inclusive of all Geomarkets (CBSA filter is set to the Philadelphia metropolitan area and Geomarket filter is set to all five Geomarkets in the Philadelphia area). Although we would like to analyze the D.C./Maryland/Virginia (DMV) area, the student list purchases we collected exclude "MD 5 -- Prince Georges Metropolitan" and "MD 7 -- Baltimore (Urban), which are substantially poorer and more Black than other Geomarkets in the DMV. Similarly, student list purchases targeting Cleveland exclude "OH 4 -- City of Cleveland (East)."

In order to isolate the impact of Geomarkets on the racial/ethnic and socioeconomic characteristics of prospect pools, we did not consider purchases that utilized student preference (e.g., campus size, intended major), low-income household, or first-generation status filters. Because nearly all orders we collected (Author, BLINDED) filter on some academic criteria, we only considered orders that included assessment threshold filters (e.g., PSAT, SAT) across a wide score range to explore whether and to what extent Geomarket contributions to prospect pools are consistent across all score ranges.

@@

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. The maps in the figure show how Chicago Geomarkets reflect the classic urban geographical structure of an inner-city core surrounded by a ring of suburbs as described by @RN4982[p. 12]. Based on the color gradient scale representing total population density, @fig-chicago-maps indicates northern suburb Geomarkets in the metropolitan area - Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie - experienced relatively little population growth from 1980 through 2020. In contrast, the most significant population increases occured in the Western Suburbs (1.2 Million to 1.6 Million) and the South and Southwest Suburbs (less than 1 Million to 1.3 Million) Geomarkets. The City of Chicago Geomarket, forming the metropolitan area's central core, was the only locality to experience a population decline, from 3.1 Million in 1980 to 2.8 Million in 2020, despite remaining the most densely populated Geomarket.

These modest population sgufts belie substantial change in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarket across the three census periods. In 1980, nearly all Geomarkets were predominantly White, with proportions ranging from 83% in the South and Southwest Suburbs to approximately 93% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 46% of its total population. By 2020, several Geomarkets that were previously predominantly White shifted to a more heterogeneous racial/ethnic distribution, including Chain of Lakes (59% White, 8% Black, 24% Hispanic; 7% Asian Pacific Islander), Northwest Suburbs (62% White, 3% Black, 18% Hispanic; 15% Asian Pacific Islander), Evanston and Skokie (64% White, 7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (56% White, 9% Black, 23% Hispanic; 9% Asian Pacific Islander), and South and Southwest Suburbs (53% White, 24% Black, 18% Hispanic; 3% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 19 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie). @fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. <!-- For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. -->

@@

From 1980 to 2020, these demographic shifts have transformed the Chicago metropolitan area's Geomarket landscape from largely a binary structure into a more tiered dynamic shaped by racial/ethnic and socioeconomic characteristics. The original Geomarket boundaries delineated significant racial and income segregation in the 1980s. This included Black, Hispanic, and Asian residents primarily in the City of Chicago Geomarket and White residents in Geomarkets forming the suburban ring of the metropolitan area. Income, poverty, and educational attainment also created a static socioeconomic divide in 1980, with southern Geormarkets (City of Chicago and South and Southwest Suburbs) characterized by lower affluence and educational attainment compared to the higher socioeconomic standing of the northwestern suburbs. However, declining White populations, Hispanic and Asian population growth, and Black out-migration, as well as some narrowing of education and poverty gaps, have contributed to a more stratified Geomarket dynamic in the metropolitan area despite significant disparities remaining. The City of Chicago Geomarket remains the critical geographical hub across all Geomarkets for racial/ethnic diversity; however, other Geomarkets include substantive numbers of residents across racial/ethnic groups (such as South and Southwest Geomarket for Black residents, Chain of Lakes for Hispanic residents, and Northwest Suburbs for Asian residents) while the the Evanston and North Shore Geomarkets remain predominantly White. While some stark socioeconomic gaps within Geomarkets have narrowed since 1980, significant geographical income segregation remains across Geomarkets despite a more clearly defined middle-to-high socioeconomic Geomarket group (Chain of Lakes, Northwest Suburbs, Western Suburbs). [I THINK SOME DETAIL CAN BE REDUCED HERE; ALSO, I THINK YOU ARE BURYING THE LEDE THAT NEEDS TO BE EMPHASIZED: THERE IS SUBSTANTIAL RACIAL AND SES INEQUALITY BETWEEN GEOMARKETS AND CHICAGO ESPECIALLY IS POOREST AND MOST NON-WHITE]

@@

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets, we analyze six orders placed by a research university targeting prospects across the entire state of Illinois. The first two orders filter on the same SAT thresholds (ranging from 1020 to 1150), the same GPA thresholds (ranging from B- to A+), and filter on the high school graduating classes of 2020 and 2021 graduating classes, respectively. The remaining four orders used similar GPA and high school class filters but differed in SAT thresholds. Two orders targeted students with relatively middle-range scores (1160 to 1300) and two orders targeted those with relatively high scores (1310 to 1600). Because all six orders included an Illinois state filter, we can utilize prospects from the Chicago metropolitan area within the resulting student lists to examine Geomarket contributions by simulating the application of a Geomarket filter across low-, middle-, and high-range SAT threshold orders.

DISCUSSION OF THIS FIGURE SOMEWHAT BURIES THE LEAD THAT CHICAGO LOOKS TOTALLY DIFFERENT THAN THE OTHERS IN TERMS OF RACIAL COMPOSITION OF PURCHASED PROFILES!
@fig-chicago-rq2-race presents the racial/ethnic composition of purchased student profiles within Geomarkets in the Chicago metropolitan area. At the lowest PSAT score thresholds (top panel, 1020-1150) the racial/ethnic distribution of prospects within Geomarkets generally reflects the demographic makeup of the metropolitan area in 2020 (see @fig-chicago-rq1-race). For example, more than 8,000 of the 16,485 total prospects in this SAT range order were concentrated in the Western Suburbs (56% White, 8% Asian, 6% Black, 26% Hispanic, 4% Multiracial) and South and Southwest Suburbs (55% White, 3% Asian, 17% Black, 22% Hispanic, and 3% Multiracial), both of which yield prospect pools comparable to the racial/ethnic makeup of their overall Geomarket populations. Similar to the larger metropolitan area, the North Shore (78% White, 8% Asian, 2% Black, 8% Hispanic) and the Northwest Suburbs (61% White, 11% Asian, 2% Black, 23% Hispanic) Geomarket pools exhibit the greatest proportions of White prospects, reflecting the Geomarket dynamics in the metropolitan area population although they yielded fewer overall students. Interestingly, while the City of Chicago is the most racially/ethnically diverse Geomarket in the larger metropolitan area, its prospect pool in @fig-chicago-rq2-race includes nearly double the proportion of Hispanic (50%) prospects relative to the larger Geomarket resident population (26% Hispanic) in @fig-chicago-rq1-race with a lower proportion of White prospects (14%).
